---
title: 'Chapter 3: Data Visualisation'
author: "Belinda Fabian"
date: "31/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

# 3.1 Introduction

R has several systems for making graphs, but ggplot2 is one of the most elegant and most versatile. ggplot2 implements the grammar of graphics, a coherent system for describing and building graphs.

ggplot2 is a package that's part of the tidyverse

data frame = rectangular collection of variables (columns) & observations (rows)

```{r load packages, echo = TRUE, results = "hide", error = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
```

# 3.2 First steps

## 3.2.1 the mpg data frame

Q: Do cars with big engines use more fuel than cars with small engines? 

Use: mpg data frame found in ggplot2 (aka `ggplot2::mpg`)
`mpg` contains observations collected by the US Environmental Protection Agency on 38 models of car.

`displ` = a car’s engine size, in litres.

`hwy` = car’s fuel efficiency on the highway, in miles per gallon (mpg)

car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance


## 3.2.2 Creating a ggplot

Create a plot of `displ` vs `hwy`

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

* Begin a plot with the function `ggplot()` --> creates a coordinate system that you can add layers to
* First argument of `ggplot()` is the dataset to use in the graph
* Complete a graph by adding one or more layers to `ggplot()` --> these are geom functions

Each geom function takes a mapping argument which defines how the variables in your dataset are mapped to visual properties

The mapping argument is **always** paired with `aes()`, and the `x` and `y` arguments of `aes()` specify which variables to map to the x and y axes --> ggplot2 looks for the mapped variables in the data argument


## 3.2.3 A graphing template

template for making graphs with ggplot2

`ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`


## 3.2.4 Exercises

1. Run `ggplot(data = mpg)`. What do you see?

```{r}
ggplot(data = mpg)
```

See a grey square as haven't told ggplot 2 which geom function or mappings to use (no data or method of visualisation)


2. How many rows are in `mpg`? How many columns?

```{r}
nrow(mpg)
ncol(mpg)
```

number of rows = 234
number of columns = 11

3. What does the `drv` variable describe? Read the help for `?mpg` to find out.

```{r}
?mpg
```

`drv` = the type of drive train (f = front-wheel drive, r = rear wheel drive, 4 = 4wd)


4. Make a scatterplot of `hwy` vs `cyl.`

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cyl, y = hwy))
```


5. What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = class, y = drv))
```

Not a useful plot as both variables are categorical


# 3.3 Aesthetic mappings

On the `displ` vs `hwy` graph above there are a few points that fall outside the negative linear relationship i.e. have a lower mileage than is expected for their engine size
--> hypothesis = these are hybrid vehicles

check the `class` for each car to find out

add a third variable to a 2D scatterplot by mapping it to an aesthestic e.g. size, shape or colour of the points

convey information about data by mapping aesthetics to variables in the data set

for example --> map the colours of points to `class` variable to see the class of each car


```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

associate the name of the aesthetic with the name of the variable inside `aes()`

scaling = when R automatically assigns a unique level of the aesthetic to each unique value of the variable

ggplot2 also adds a legend explaining which levels correspond to which values

turns out the outliers in `displ` vs `hwy` are two-seater cars i.e. sports cars


could also map class to size aesthetic --> when we do this will get an error as we are mapping an unordered variable (`class`) to an ordered aesthetic (`size`) - R will still do it, but it's not a good idea as it implies something about the data that is not accurate)

could also map to alpha which controls transparency of point or shape which controls shape of points



```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

use `aes()` to associate the name of the aesthetic with a variable to display

once aesthetic is mapped, ggplot2 takes care of the rest - it selects a reasonable scale, and it constructs a legend that explains the mapping between levels and values. For x and y aesthetics it creates an axis line with tick marks and a label.

can map an aesthetic manually - this is outside of `aes()`

for example, make all points in the plot blue

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), colour = "blue")
```

use a level that makes sense for that aesthetic:
* The name of a color as a character string
* The size of a point in mm
* The shape of a point as a number from 1-25 (run `vignette("ggplot2-specs")` to see shapes)


## 3.3.1 Exercises

1. What’s gone wrong with this code? Why are the points not blue?

To manually assign an aesthetic it has to be outside the `aes()` bracket
```{r}
# incorrect code
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))

# corrected code
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```


2. Which variables in `mpg` are categorical? Which variables are continuous? (Hint: type `?mpg` to read the documentation for the dataset). How can you see this information when you run `mpg`?

can see this information from the data class displayed underneath each variable name/column name

categorical variables =  manufacturer, model, displ, year, cyl, trans, drv, fl, class

continuous variables = cty, hwy

```{r}
?mpg
str(mpg)
mpg
```

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

When colour is mapped to a continuous variable a colour scale is used rather than discrete colours

Some aesthetics (i.e. shape) can't be mapped to a continuous variable

Some aesthetics have problems when mapping to a categorical variable with large number of levels e.g. more than 6 shapes

Some aesthetics aren't great for differentiating when there are many levels of the aesthetic, regardless of continuous/categorical variables

```{r}
# map aesthetics to categorical variable
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = manufacturer))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = manufacturer))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = manufacturer))

# map aesthetics to continuous variable
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = cty))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = cty))

# ggplot(data = mpg) +
#   geom_point(mapping = aes(x = displ, y = hwy, shape = cty))

```


4. What happens if you map the same variable to multiple aesthetics?

Information is conveyed in two ways, now there are two legends - makes plot harder to understand, unnecessary

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x= displ, y = hwy, colour = cyl, size = cyl))
```

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

the stroke aesthetic determines the thickness of the line around a point - works with all shapes

shapes 21-24 have stroke `colour` and `fill` - size of the filled part controlled by `size`, size of the stroke controlled by `stroke` (both measured in mm)

```{r}
?geom_point

vignette("ggplot2-specs")

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), shape = 22, fill = 5, stroke = 1, size = 3)

```

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

Splits the data into true and false for those values < 5 and > 5, respectively, and colours the points based on true and false values

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = manufacturer, y = hwy, colour = displ < 5))
```



# 3.4 Common problems

* unmatched parentheses
* unmatched quotation marks
* `+` in console means unfinished expression - either complete the expression or press ESC
* `+` in ggplot2 expressions go at the end of the line
* get help by running `?function_name`
* check error messages - google to find help


# 3.5 Facets

Facetting splits plots into subplots that display one subset of the data
Use discrete variables for faceting

Facet by single variable using `facet_wrap(~ variable)`

```{r}
ggplot(data = mpg) +
  geom_point( mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ cyl, nrow = 2)
```

Facet using combination of two variables using `facet_grid(variable1 ~ variable2)` 
--> in this example variable1 will be in the rows and variable2 will be in the columns

If don't want to facet in rows or colums dimension use `.` instead of variable name e.g. `facet_grid(. ~ cyl)` 
This forces the facet to be in rows/columns rather than a default grid from `facet_wrap()`

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)

ggplot(data = mpg) +
  geom_point(mapping = aes(x= displ, y = hwy))+
  facet_grid(. ~ cyl)
```


## 3.5.1 Exercises

1. What happens if you facet on a continuous variable?

Facet functions will still work, but will treat each of the continuous data points as discrete so there will be many subplots

```{r}
names(mpg)

head(mpg, n = 20)

mpg %>% 
  count(cty)

# facet using continuous variable `cty`
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ cty)
```


2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

In the first plot the empty cells are where there is no data for that combination of facetted variables i.e. there is no data that has the combination of cyl = 5 and drv = 4

The second plot is of cyl and drv directly - can see that there is no data point of cyl = 5 and drv = 4


```{r}
# original plot with empty cells from face_grid()
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)

# new plot directly mapping drv and cyl to axes
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```



3. What plots does the following code make? What does . do?

First plot facets data by `drv` in rows (y-axis)
Second plot facets data by `cyl` in columns (x-axis)

The use of `.` here ignores the dimension for facetting where it is used

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```


4. Take the first faceted plot in this section:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```


What are the advantages to using faceting instead of the colour aesthetic? 
- colour blind friendly
- easy to see large differences in data
- colours can have small visual distance between them making them hard to distinguish
- can change the scale for some facets if needed

What are the disadvantages? 
- hard to see small differences in data
- use of different scales makes it hard to compare data points across facets

How might the balance change if you had a larger dataset?
- lots of data points might make the data overlap more, so difficult to see the differences in the data using the colour aesthetic --> in this case would be better to use facetting


5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

nrow and ncol prescribe the number of rows or columns used in the facet wrap

as.table also controls the layout of the panels

facet_grid doesn't have nrow and ncol arguments as the number of rows and columns is determined by the number of unique values in the variables used


6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?

To take advantage of screen layouts being wider than they are tall


# 3.6 Geometric objects

geom is a visual/geometric object that a plot uses to represent data

to change the geom in a plot, change the geom function added to `ggplot()`

every geom function take a `mapping` argument, but not every aesthetic works in ebery geom 
e.g. can't set the shape of a line

ggplot2 has over 40 geoms, plus more in extension packages

many geoms use a single geometric object to display multiple rows of data - set the group 
aesthetic to a categorical variable to draw multiple objects (one for each unique value of the 
grouping variable) - using group aesthetic doesn't add legend or distinguishing features to geoms

mapping an aesthetic to a discrete variable will automatically group the data - automatically adds 
legend and distinguishing features to geoms

to display multiple geoms in the same plot add multiple geoms functions to `ggplot()`
this can introduce duplication in the code - if all geoms have the same mappings can move them into
the `ggplot()` function - this will apply these mappings as global settings that apply to all geoms

if mappings are in a geom function, ggplot2 will treat them as local mappings for the layer
will use these mappings to extend or overwrite the global mappings for that layer only
this makes it possible to display different aesthetics in different layers

can specify different data for each layer - for example, adding a local data argument in geom_smooth() will override the global data argument in ggplot() for that layer only


## 3.6.1 Exercises

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

line chart = `geom_line()`
boxplot = `geom_boxplot()`
histogram = `geom_histogram()`
area chart = `geom_area()`


2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

Prediction: plot with displ on x axis, hqy on y axis, points and lines coloured by drv, legend will be present with colours of drv, no standard error shading around lines

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```


3. What does show.legend = FALSE do? What happens if you remove it?
Why do you think I used it earlier in the chapter?

`show.legend()` relates to presence/absence of legend on plot, if remove this code then the legend will be present (default is to show legend on plot)


4. What does the se argument to geom_smooth() do?

the se argument relates to the presence/absence of the standard error shading around a fit line


5. Will these two graphs look different? Why/why not?

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

These plots will look the same. putting the same data and mappings in each of the geoms is the same as putting the data and mappings once inside the `ggplot()` function.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```



6. Recreate the R code necessary to generate the following graphs.

```{r}

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(mapping = aes(group = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = drv)) +
  geom_smooth(se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = drv)) +
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(fill = drv), shape = 21, colour = 'white', stroke = 1)

```





# 3.7 Statistical transformations

## 3.7.1 Exercises


# 3.8 Position adjustments

## 3.8.1 Position adjustments


# 3.9 Coordinate systems

## 3.9.1 Exercises


# 3.10 The layered grammar of graphics





