---
title: 'Chapter 4: Workflow - basics'
author: "Belinda Fabian"
date: "04/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

# 4.1 Coding basics

type straight in the console to use R as a calculator

use `<-` to assign a value to an object

# 4.2 What's in a name?

object names must start with a letter, and can only contain letters, numbers, `_` and `.`

inspect an object by typing its name

use `tab` for RStudio’s completion facility (autocomplete object names)

Typos matter. Case matters.


# 4.3 Calling functions

can use tab autocompletion with the start of function names

if there is incomplete infomration and R can't execute a function then will give a `+` symbol in the console - asking for more information to be entered
either enter the needed information or press `ESC` to abort and return to the prompt

surround an assignment with parentheses to print the result of an assignment to the console


# 4.4 Exercises

1. Why does this code not work?

my_variable <- 10  
my_varıable  
> Error in eval(expr, envir, enclos): object 'my_varıable' not found

Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)

Error in spelling of object name in second line


2. Tweak each of the following R commands so that they run correctly:

library(tidyverse)

ggplot(dota = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

fliter(mpg, cyl = 8)
filter(diamond, carat > 3)

```{r load packages, echo = TRUE, results = "hide", error = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
```

```{r}
names(mpg) # error is about name 'displ' - check what column names are

ggplot(data = mpg) + # error is in spelling of 'data'
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8) # corrected spelling of filter and changed = to == 

data() # object diamond not found, check included data set names to find correct spelling

filter(diamonds, carat > 3) # correct data set name is 'diamonds'
```


3. Press Alt + Shift + K. What happens? How can you get to the same place using the menus?

Brings up a list of keyboard shortcuts
Use Tools menu to bring up this list

