---
title: "Chapter 5: Data Transformation"
author: "Belinda Fabian"
date: "03/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

# 5.1 Introduction

transforming data using the dplyr package and dataset on flights departing New York City in 2013

## 5.1.1 Prerequisites

```{r load packages, echo = TRUE, results = "hide", error = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(nycflights13)
```

message that’s printed when load the tidyverse - dplyr overwrites some functions in base R

## 5.1.2 nycflights13

explore the basic data manipulation verbs of dplyr using `nycflights13::flights`

data frame contains all 336,776 flights that departed from New York City in 2013
data comes from the US Bureau of Transportation Statistics, and is documented in `?flights`

this datafram is a tibble --> shows the first few rows and all the columns that fit on one screen
Tibbles are data frames, but slightly tweaked to work better in the tidyverse
more detail in chapter 10
```{r}
nycflights13::flights
```

row of 3 (or 4) letter abbreviations under column names describes the type of each variable:
*`int` stands for integers
*`dbl` stands for doubles, or real numbers
*`chr` stands for character vectors, or strings
*`dttm` stands for date-times (a date + a time)

three other common types of variables will encounter later:
*`lgl` stands for logical, vectors that contain only TRUE or FALSE
*`fctr` stands for factors, R uses to represent categorical variables with fixed possible values
*`date` stands for dates

## 5.1.3 dplyr basics

5 key dplyr functions that can solve the vast majority of data manipulation challenges:
*Pick observations by their values (`filter()`)
*Reorder the rows (`arrange()`)
*Pick variables by their names (`select()`)
*Create new variables with functions of existing variables (`mutate()`)
*Collapse many values down to a single summary (`summarise()`)

These can all be used in conjunction with `group_by()` which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. These six functions provide the verbs for a language of data manipulation.

All verbs work similarly:
*The first argument is a data frame
*The subsequent arguments describe what to do with the data frame, using variable names (no quotes)
*The result is a new data frame

These properties make it easy to chain together multiple simple steps to achieve a complex result


# 5.2 Filter rows with `filter()`

## 5.2.1 Comparisons

## 5.2.2 Logical operators

## 5.2.3 Missing values

## 5.2.4 Exercises


# 5.3 Arrange rows with `arrange()`

## 5.3.1 Exercises


# 5.4 Select columns with `select()`

## 5.4.1 Exercises


# 5.5 Add new variables with `mutate()`

## 5.5.1 Useful creation functions

## 5.5.2 Exercises


# 5.6 Grouped summaries with `summarise()`

## 5.6.1 Combining multiple operations with pipe

## 5.6.2 Missing values

## 5.6.3 Counts

## 5.6.4 Useful summary functions

## 5.6.5 Grouping by muliple variables

## 5.6.6 Ungrouping

## 5.6.7 Exercises


# 5.7 Grouped mutates (and filters) 

## 5.7.1 Exercises


