---
title: "Chapter 5: Data Transformation"
author: "Belinda Fabian"
date: "03/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

```{r load packages}
library(tidyverse)
```

# 5.1 Introduction

## 5.1.1 Prerequisites

## 5.1.2 nycflights13

## 5.1.3 dplyr basics


# 5.2 Filter rows with `filter()`

## 5.2.1 Comparisons

## 5.2.2 Logical operators

## 5.2.3 Missing values

## 5.2.4 Exercises


# 5.3 Arrange rows with `arrange()`

## 5.3.1 Exercises


# 5.4 Select columns with `select()`

## 5.4.1 Exercises


# 5.5 Add new variables with `mutate()`

## 5.5.1 Useful creation functions

## 5.5.2 Exercises


# 5.6 Grouped summaries with `summarise()`

## 5.6.1 Combining multiple operations with pipe

## 5.6.2 Missing values

## 5.6.3 Counts

## 5.6.4 Useful summary functions

## 5.6.5 Grouping by muliple variables

## 5.6.6 Ungrouping

## 5.6.7 Exercises


# 5.7 Grouped mutates (and filters) 

## 5.7.1 Exercises


