---
title: 'Chapter 7: Exploratory Data Analysis'
author: "Belinda Fabian"
date: "16/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

# 7.1 Introduction

This chapter is about using visualisation and transformation to explore data in a systematic way, a task that statisticians call exploratory data analysis (EDA)

EDA is an iterative cycle
*Generate questions about your data.
*Search for answers by visualising, transforming, and modelling your data.
* Use what you learn to refine your questions and/or generate new questions.

EDA is not a formal process with a strict set of rules, EDA is a state of mind
During initial phases of EDA feel free to investigate every idea 

EDA is an important part of any data analysis, even if the questions are already set - always need to investigate  quality of data
Data cleaning is just one application of EDA: ask questions about whether  data meets expectations 
To do data cleaning, need to deploy all tools of EDA: visualisation, transformation, and modelling

In this chapter: combine dplyr and ggplot2 to interactively ask questions, answer them with data, and ask new questions

```{r load packages}
library(tidyverse)
```


# 7.2 Questions



# 7.3 Variation

## 7.3.1 Visualising distributions

## 7.3.2 Typical values

## 7.3.3 Unusual values

## 7.3.4 Exercises



# 7.4 Missing values

## 7.4.1 Exercises



# 7.5 Covariation

## 7.5.1 A categorical and continuous variable

### 7.5.1.1 Exercises

## 7.5.2 Two categorical variables

### 7.5.2.1 Exercises

## 7.5.3 Two continuous variables

#### 7.5.3.1 Exercises



# 7.6 Patterns & models



# 7.7 ggplot2 calls



# 7.8 Learning more

Learn more about the mechanics of ggplot2 in the ggplot2 book: https://amzn.com/331924275X
includes dplyr and tidyr code & explores all the facets of visualisation
thru uni likely can get electronic version free through SpringerLink

Another useful resource: R Graphics Cookbook by Winston Chang - uuch of the content is available online at http://www.cookbook-r.com/Graphs/

Graphical Data Analysis with R, by Antony Unwin
Book-length treatment similar to the material covered in this chapter, but has the space to go into much greater depth.
