---
title: "RStudio tidyverse certification practice test"
author: "Belinda Fabian"
date: "02/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# persistently change working directory for all code chunks to the project working directory
```

```{r load packages, echo = TRUE, results = "hide", error = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
```


# RStudio tidyverse certification practice test
https://education.rstudio.com/blog/2020/02/instructor-certification-exams/

## Question 4

The file ranking.csv contains two columns:

The ID of an item being rated.
A rating, which is one of “negative”, “positive”, “indifferent”, or “wtf” (meaning the respondent didn’t understand the question).
There are multiple ratings for each item.

The plot below shows this data (see plot at https://education.rstudio.com/blog/2020/02/instructor-certification-exams/)

Each dot represents one item i.
The size of the circles shows the total number of ratings for item i.
The X coordinate for item i is the percentage of ratings for that item that are “negative”.
The Y coordinate for item i is the percentage of ratings for that item that are “positive”.
The regression line is created using the ‘lm’ method.

Re-create this plot using the tidyverse and ggplot2, fixing any mistakes you notice along the way.


```{r read data}
# import data
ranking <- read_csv("data_raw/ranking.csv")

#original data
ranking
```

```{r}

# my answer

ranking_adj <- ranking %>% 
  count(rank, item) %>%
  pivot_wider(names_from = rank, values_from = n) %>% 
  mutate_at(vars(positive, negative, indifferent, wtf), replace_na, 0) %>% 
  mutate(percent_positive = positive/(positive + negative + indifferent + wtf)) %>% 
  mutate(percent_negative = negative/(positive + negative + indifferent + wtf)) %>% 
  mutate(total_rankings = positive + negative + indifferent + wtf)
 
ranking_adj
```

```{r}
ggplot(data = ranking_adj) +
  geom_point(mapping = aes(x = percent_negative, 
                           y = percent_positive, 
                           size = total_rankings), 
             alpha = 0.5) +
  geom_smooth(mapping = aes(x = percent_negative, 
                           y = percent_positive), 
              method = "lm") +
  labs(x = "negative", y = "positive")
```

