---
title: "Study group questions"
author: "Belinda Fabian"
date: "16/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

Jen Richmond's questions & solutions available here: https://github.com/jenrichmond/r4ds_exercises

# Chapter 5: Data Transformation

Refer to notes in '5_Data_Transformation.Rmd'

```{r load packages, echo = TRUE, results = "hide", error = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(nycflights13)
library(lubridate)
```

## 1. Question from Belinda Fabian

Using the nycflights13::flights dataframe

Does the flight delay increase over the course of a day (i.e. is the arrival delay on the first flight of the day less than the arrival delay at the end of the day)?

  Yes, delays propogate throughout the day

Does this difference vary by month, airline or origin?

  Delays increase the most over the course of the day in July , delays increase the least over the course of the day in October 
  
  Lots of variation in arrival delays over the course of the day between carriers --> some carriers only have small number of flights for each departure time, so distorts the results
  Flights on FL are most delayed by the end of the day, flights on B6 are the least delayed by the end of the day
  
  Flights originating at JFK are the least delayed by the end of the day
  
  B6 flights out of JFK are least delayed

```{r}
names(flights)

# exclude flights before 5am as these are generally delayed from the day before
filtered_flights <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_delay)) %>% 
  filter(dep_time > 500) 

# does the flight delay increase over the course of a day? --> combine all days
filtered_flights %>% 
  group_by(dep_time) %>% 
  summarise(mean_arr_delay = mean(arr_delay)) %>% 
  ggplot(mapping = aes(x = dep_time, y = mean_arr_delay)) +
  geom_point()

# does the difference in the flight delay over the day vary by month?
filtered_flights %>% 
  group_by(dep_time, month) %>% 
  summarise(mean_arr_delay = mean(arr_delay)) %>% 
  ggplot(mapping = aes(x = dep_time, y = mean_arr_delay)) +
  geom_line() +
  facet_wrap(~month)

filtered_flights %>% 
  group_by(month, dep_time) %>% 
  summarise(mean_arr_delay = mean(arr_delay)) %>% 
  filter(mean_arr_delay == max(mean_arr_delay)) %>% 
  ungroup() %>% 
  slice_min(mean_arr_delay)

# does the difference in the flight delay over the day vary by airline?
filtered_flights %>% 
  group_by(carrier, dep_time) %>% 
  summarise(mean_arr_delay = mean(arr_delay)) %>% 
  ggplot(mapping = aes(x = dep_time, y = mean_arr_delay)) +
  geom_line() +
  facet_wrap(~carrier)

## check on strange results for HA
# filtered_flights %>% 
#   filter(carrier == "HA") %>% 
#   group_by(dep_time) %>% 
#   count()

# does the difference in the flight delay over the day vary by origin?
filtered_flights %>% 
  group_by(origin, dep_time) %>% 
  summarise(mean_arr_delay = mean(arr_delay)) %>% 
  ggplot(mapping = aes(x = dep_time, y = mean_arr_delay)) +
  geom_line() +
  facet_wrap(~origin)

# interaction in  flight delay between airline and origin?
filtered_flights %>% 
  group_by(origin, carrier, dep_time) %>% 
  summarise(mean_arr_delay = mean(arr_delay)) %>% 
  ggplot(mapping = aes(x = dep_time, y = mean_arr_delay)) +
  geom_line() +
  facet_grid(origin~carrier)

```

## 2. Question from Jen Richmond

This data comes from: Tidy Tuesday Week 39 2020 https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md

```{r get data}
peaks <- download.file("https://raw.githubusercontent.com/jenrichmond/r4ds_exercises/main/data/peaks.csv", destfile = "data_raw/peaks.csv")

peaks <- read_csv("data_raw/peaks.csv")
```

### 1. How many different himalayan peaks are represented in this dataset?

  468 different peaks

```{r}
names(peaks)

# using pipe and count
peaks %>% 
  count(n_distinct(peak_id, na.rm = TRUE))

# simplified
n_distinct(peaks$peak_id)

```

### 2. what proportion of them have been climbed vs not?

  73% of peaks have been climbed

```{r}
peaks %>% 
  summarise(propn_climbed = mean(climbing_status == "Climbed"))
```

### 3. is there any difference in the average height of those that have been climbed vs not?

  Average height of the peaks climbed is greater than the average height of the unclimed peaks

```{r}
peaks %>% 
  group_by(climbing_status) %>%
  summarise(mean_height = mean(height_metres))

```

### 4. Which peak was climbed first?

  Peak Sharphu II was climbed first in the year 201 --> this may be a data entry error
  Peak Langpo was climbed second in 1909
  
```{r}
peaks %>% 
  slice_min(first_ascent_year, n = 2)
```

### 5. How many of the first ascents were achieved by teams from Australia or New Zealand? 

  7 peaks were first ascended by a team soley from Australia or New Zealand

```{r}
names(peaks)
head(peaks)

# check how the observations are listed in first_ascent_country
peaks %>% 
  group_by(first_ascent_country) %>% 
  count(first_ascent_country)

# peaks first ascended by only Australians or New Zealanders
peaks %>% 
  filter(first_ascent_country %in% c("Australia", "New Zealand"))

```

### 6. How many of first ascents had teams that involved climbers from NZ or Australia (as well as people from other countries)?

    Teams from Australia or New Zealand were present for the first ascent of 18 peaks
    
```{r}
# peaks first ascended with Australians or New Zealanders present
peaks %>% 
  filter(str_detect(first_ascent_country, "Australia|Zealand"))
```

### 7. Add a new variable that lists TRUE/FALSE re whether or not there was a Kiwi on the climbing team. Use it to calc what proportion of first ascents had Kiwis on the team?

  4% of first ascent teams included Kiwis

```{r}
peaks %>% 
  mutate(kiwi = str_detect(first_ascent_country, "Zealand")) %>%
  summarise(propn_with_kiwi = mean(kiwi, na.rm = TRUE))
```

### 8. How many of the peaks climbed, do we not know the date/country of the ascent?

  For 6 peaks that have been climbed there is no information on date or country of first ascent

```{r}
names(peaks)

peaks %>%
  filter(climbing_status == "Climbed") %>% 
  filter(is.na(first_ascent_year) | is.na(first_ascent_country))

peaks %>%
  group_by(climbing_status) %>%
  count(is.na(first_ascent_year))
```

### 9. Which peak is closest in height to everest?

  Kangchenjunga is the second highest peak

```{r}
peaks %>% 
  slice_max(height_metres, n = 2)
```

### 10. What proportion of the peaks taller than 7000 ft have been climbed? 

  88% of the peaks taller than 7000 feet have been climbed

```{r}
peaks %>% 
  filter(height_metres > 7000) %>% 
  summarise(propn_climbed = mean(climbing_status == "Climbed"))
```



# Chapter 7: Exploratory Data Analysis

Refer to notes in '7_Exploratory_Data_Analysis.Rmd'

## 1. Question from Belinda Fabian


```{r}

```

## 2. Question from Jen Richmond


```{r}

```


## 3. Question from Paula Andrea


```{r}

```


## 4. Question from Disha Patel


```{r}

```








# Template for inserting questions

## 1. Question from Belinda Fabian


```{r}

```

## 2. Question from Jen Richmond


```{r}

```


## 3. Question from Paula Andrea


```{r}

```


## 4. Question from Disha Patel


```{r}

```


