---
title: "Study Group Questions"
author: "Belinda Fabian"
date: "16/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..") 
# change working directory for all code chunks to the project working directory
```

# Overview of study group

Study group for RStudio tidyverse certification includes:
- Belinda Fabian (@beacurious)
- Jen Richmond (@)
- Paula Andrea (@)
- Disha Patel (@)

Working through 'R for Data Science' by Hadley Wickham & Garrett Grolemund
https://r4ds.had.co.nz

Typically covering 1 chapter per fortnight in own private study
Meeting fortnightly - each person writes a  question about that fortnight's content for the other participants to answer
Questions shared on Slack (conversation in ROpenSci slack workspace) or GitHub repo

Creator noted for each question

# Chapter 5: Data Transformation

1. Question from Belinda Fabian

Does the flight delay increase over the course of a day (i.e. is the arrival delay on the first flight of the day less than the arrival delay at the end of the day)?

Does this difference vary by month, airline or departure city?

```{r}

```

2. Question from Jen Richmond


```{r}

```


3. Question from Paula Andrea


```{r}

```


4. Question from Disha Patel


```{r}

```


# Template for inserting questions

1. Question from Belinda Fabian


```{r}

```

2. Question from Jen Richmond


```{r}

```


3. Question from Paula Andrea


```{r}

```


4. Question from Disha Patel


```{r}

```


